---
title: "Update 06.03.2019"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(foreign)
library(survival)
library(broom)
library(stargazer)
library(tidyverse)
library(data.table)
library(TraMineR)
library(reshape)
library("RColorBrewer")
library(cluster)

# 0.3 Set right working directory
dir()
setwd("C:/Users/y4956294S/Documents/LONGPOP/Subproject 2 - SE differences in transition to dependency/R code/EDAD")

# load traminer data to show graphs
load("datasets/020_traMay_D13.RData")
tra_may_C_50 <- tra_may_D13 %>% filter(EdadInicioDisca13>=50) %>% filter(Edadinicio_cuidado<100) %>% filter(!is.na(EdadInicioDisca13)) %>% 
 filter(DISCA13_AGE<100) %>% filter(EdadInicioDisca13 <= EDAD) 
tra_may_M <- tra_may_C_50 %>% filter(SEXO=="Varón") %>% mutate(SEXO="Male")
tra_may_F <- tra_may_C_50 %>% filter(SEXO=="Mujer") %>% mutate(SEXO="Female")
seqmat_M <- tra_may_M[,c(10:60)]
seqmat_F <- tra_may_F[,c(10:60)]

SeqAlphab_C <- c("DF", "ID", "DC")
# Define color scheme
Brewer_C <- brewer.pal(3, "Set1")

# Males
DisSeq_M <- seqdef(seqmat_M,informat = "STS", alphabet = SeqAlphab_C, id="auto", cpal =  Brewer_C ,start = 55, 
                   labels = c("Disability Free", "Idependent","Care"))

# Females
DisSeq_F <- seqdef(seqmat_F,informat = "STS", alphabet = SeqAlphab_C, id="auto", cpal =  Brewer_C ,start = 55, 
                   labels = c("Disability Free", "Idependent","Care"))

load("datasets/020_traMay_D13M.RData")
load("datasets/020_traMay_D13F.RData")

# Disca 13 - 50+
load(file = 'datasets/030_linkmay_M_50ADL.RData')
load(file = 'datasets/030_linkmay_F_50ADL.RData')

```

# Analysing Sequences to cluster pathways - What is new?
## Clustering
There are a dozens of dissimilarity measures to group individual sequences, many of them more precise than classic OM. Refering to [Studer and Ritschard (2016)](https://rss.onlinelibrary.wiley.com/doi/pdf/10.1111/rssa.12125), I chose to use an OM mechanism sensitive to both the spell length and the timing of sequences (OMspell).


## Population
The population used is a little different for two reasons. First, as I discussed with Antonio, I only used individuals with an onset of disability after age 50 to capture exclusively old-age disability. Second, I found some problems in the matching process that led to unreasonable age distribution (Age of onset of disability after age at survey). As I could not fix all the issues and we still expect the updated data, I deleted the rest of the cases listwise. We end up with `r length(link.may_F$Id)` females and `r length(link.may_M$Id)` male individuals.

## Defining Disability
The onset of disability is defined as age at which the first of the following 13 activities (of daily living) became difficult/was impossible in consequence of a disability or disease.

1. Changing the body posture
1. Walking and moving inside the house
1. Walking and moving outside
1. Sitting down and using public transport
1. Wash and dry different body parts
1. Basic hygene
1. Urination
1. Going to the toilet
1. To dress and undress
1. Eating and drinking
1. Shopping (groceries)
1. Preparing food/cooking
1. Household task (cleaning the house)

# Analysis of mortality and survival
## Descriptive
Since the optimal group/cluster size in all analyses of sequences (choosing different disability indicators) was **two**, we pre-defined two distinct pathways for each sex. See the graphs for men (first graph) and women (second graph).
```{r cluster males, echo=FALSE}
# males 2 and 4 groups
seqdplot(DisSeq_M, group=tra_may_M$cluster2)
seqdplot(DisSeq_F, group=tra_may_F$cluster2)
```
The bars in the following graphs represent mean time in each state by cluster. To keep the same order, men first and then females.
```{r meantime males, echo=FALSE}
# males 2 and 4 groups
seqmtplot(DisSeq_M, group=tra_may_M$cluster2)
seqmtplot(DisSeq_F, group=tra_may_F$cluster2)
```

The following tables are median at death and mean ages at death the men and women separated.
```{r agedeath, echo=FALSE}
# males 2 clusters
link.may_M %>% group_by(cluster2) %>% summarize(mean=mean(age.ex[event==1]), median=median(age.ex[event==1]))
# females 2 cluster
link.may_F %>% group_by(cluster2) %>% summarize(mean=mean(age.ex[event==1]), median=median(age.ex[event==1]))
```
Purely looking at relative distributions, these are durations spent in disability (according to our definition) for all individuals grouped by those who experience the event and those who didn´t.
```{r disdis, echo=FALSE}
# Males
link.may_M <- link.may_M %>% mutate(dur_dis = round(age.ex,0)-round(EdadInicioDisca13,0))
# hist(link.may_M$dur_dis)
#summary(link.may_M$dur_dis)
# event distribution by duration in disability
# --------------------------------------------
DUR_Plot_M <- link.may_M %>% mutate(event = as.factor(event)) %>% 
  ggplot(aes(x=dur_dis, fill=event)) +
  geom_histogram(aes(y=0.5*..density..), binwidth=0.5) +
  scale_x_continuous(name = "Duration in Disability (Years)") +
  scale_y_continuous(name = "Relative Frequency",labels = scales::percent) +
  scale_fill_manual(name = "", values=c("#000000", "#A9A9A9"), labels = c("Survivors", "Deceased")) +
  theme_bw()
DUR_Plot_M <- DUR_Plot_M + theme(legend.position = c(0.85, 0.80)) + 
  theme(axis.text=element_text(size=12), axis.title=element_text(size=12,face="bold"))
 
DUR_Plot_M
                                                                          
link.may_F <- link.may_F %>% mutate(dur_dis = round(age.ex,0)-round(EdadInicioDisca13,0))
# hist(link.may_F$dur_dis)
# summary(link.may_F$dur_dis)

# event distribution by duration in disability
# --------------------------------------------
DUR_Plot_F <- link.may_F %>% mutate(event = as.factor(event)) %>% 
  ggplot(aes(x=dur_dis, fill=event)) +
  geom_histogram(aes(y=0.5*..density..), binwidth=0.5) +
  scale_x_continuous(name = "Duration in Disability (Years)") +
  scale_y_continuous(name = "Relative Frequency",labels = scales::percent) +
  scale_fill_manual(name = "", values=c("#000000", "#A9A9A9"), labels = c("Survivors", "Deceased")) +
  theme_bw()
DUR_Plot_F <- DUR_Plot_F + theme(legend.position = c(0.85, 0.80)) + theme(axis.text=element_text(size=12),        axis.title=element_text(size=12,face="bold"))

DUR_Plot_F
```

## KME Plots

# Age as time axis or not?
Here we will have an important discussion point. As expected, individuals with a late onset of disability have higher survival probabilities at every single age. Interesting is the difference between men and women (area between the two graphs).
However, these kind of KME don´t make much sense as the curves don´t start at the same point in time and with a risk score lower than 1.

```{r KME1, echo=FALSE}
mfit.1a <- survfit(coxph(Surv(time=EDAD,
                              time2 = age.ex,
                              event = event) ~ 1, data = subset(link.may_M, cluster2=="early onset")), data = subset(link.may_M, cluster2=="early onset"),
                   type = "kaplan-meier")
mfit.1b <- survfit(coxph(Surv(time=EDAD,
                              time2 = age.ex,
                              event = event) ~ 1, data = subset(link.may_M, cluster2=="late onset")), data = subset(link.may_M, cluster2=="late onset"),
                   type = "kaplan-meier")

mfit.1c <- survfit(coxph(Surv(time=EDAD,
                              time2 = age.ex,
                              event = event) ~ 1, data = subset(link.may_F, cluster2=="early onset long")), data = subset(link.may_F, cluster2=="early onset long"),
                   type = "kaplan-meier")

mfit.1d <- survfit(coxph(Surv(time=EDAD,
                              time2 = age.ex,
                              event = event) ~ 1, data = subset(link.may_F, cluster2=="late onset short")), data = subset(link.may_F, cluster2=="late onset short"),
                   type = "kaplan-meier")

KME.Clusta <- tidy(mfit.1a) %>% select(estimate, time) %>% mutate(disab = "early") %>% mutate(sex="male")
KME.Clustb <- tidy(mfit.1b) %>% select(estimate, time) %>% mutate(disab = "late") %>% mutate(sex="male")
KME.Clustc <- tidy(mfit.1c) %>% select(estimate, time) %>% mutate(disab = "early") %>% mutate(sex="female")
KME.Clustd <- tidy(mfit.1d) %>% select(estimate, time) %>% mutate(disab = "late") %>% mutate(sex="female")

KME.CLustM <- union(KME.Clusta, KME.Clustb) %>% union(KME.Clustc) %>% union(KME.Clustd)
KME1 <- KME.CLustM %>% ggplot() +
  geom_step(aes(x=time, y=estimate, color=disab)) +
  scale_y_continuous(name = "Survival Probability")                  +
  scale_x_continuous(name = "Age") +
  scale_colour_manual(values = c("orange", "darkgrey"), name="")     +
  theme_bw()
KME1 + facet_grid(. ~ sex) + theme(legend.position = c(0.85, 0.80)) + theme(axis.text=element_text(size=12),        axis.title=element_text(size=12,face="bold"))
```

Here just for men, the option with more than 2 groups.
```{r KME2, echo=FALSE}

## Males
# exit time (month/12+year)
link.may_M <- link.may_M %>% mutate(t.salida = (a_salida+(m_salida/12)) - 2006.99)
#hist(link.may_M$t.salida)
# entry time (month/12+year) - onset of disability
# link.may_M <- link.may_M %>% mutate(t.entrada =  (EDAD-DISCA13_AGE)) # same story with inicio Disca44
# hist(link.may_M$t.entrada)


## Females
# exit time (month/12+year)
link.may_F <- link.may_F %>% mutate(t.salida = (a_salida+(m_salida/12)) - 2006.99)
#hist(link.may_F$t.salida)
# entry time (month/12+year) - onset of disability
# link.may_F <- link.may_F %>% mutate(t.entrada = (EDAD-DISCA13_AGE)) # same story with inicio Disca44
# hist(link.may_F$t.entrada)


mfit.2a <- survfit(coxph(Surv(time = t.salida,
                              event = event) ~ 1, data = subset(link.may_M, cluster2=="early onset")), data = subset(link.may_M, cluster2=="early onset"),
                   type = "kaplan-meier")
mfit.2b <- survfit(coxph(Surv(time=t.salida,
                              event = event) ~ 1, data = subset(link.may_M, cluster2=="late onset")), data = subset(link.may_M, cluster2=="late onset"),
                   type = "kaplan-meier")

mfit.2c <- survfit(coxph(Surv(time= t.salida,
                              event = event) ~ 1, data = subset(link.may_F, cluster2=="early onset long")), data = subset(link.may_F, cluster2=="early onset long"),
                   type = "kaplan-meier")

mfit.2d <- survfit(coxph(Surv(time= t.salida,
                              event = event) ~ 1, data = subset(link.may_F, cluster2=="late onset short")), data = subset(link.may_F, cluster2=="late onset short"),
                   type = "kaplan-meier")

KME.Clusta <- tidy(mfit.2a) %>% select(estimate, time) %>% mutate(disab = "early") %>% mutate(sex="male")
KME.Clustb <- tidy(mfit.2b) %>% select(estimate, time) %>% mutate(disab = "late") %>% mutate(sex="male")
KME.Clustc <- tidy(mfit.2c) %>% select(estimate, time) %>% mutate(disab = "early") %>% mutate(sex="female")
KME.Clustd <- tidy(mfit.2d) %>% select(estimate, time) %>% mutate(disab = "late") %>% mutate(sex="female")

KME.CLustM <- union(KME.Clusta, KME.Clustb) %>% union(KME.Clustc) %>% union(KME.Clustd)
KME2 <- KME.CLustM %>% ggplot() +
  geom_step(aes(x=time, y=estimate, color=disab)) +
  scale_y_continuous(name = "Survival Probability")                  +
  scale_x_continuous(name = "Years since survey") +
  scale_colour_manual(values = c("orange", "darkgrey"), name="")     +
  theme_bw()
KME2 + facet_grid(. ~ sex) + theme(legend.position = c(0.85, 0.80)) + theme(axis.text=element_text(size=12),        axis.title=element_text(size=12,face="bold"))

```


## Regression
Following are different survival models (mostly Cox models). Still on the table to discuss using a parametric model (might be more accurate for adult mortality, might be not).
The models below use time between survey data and date of death/censorship as time dimensions. In other words, we compare the survival experience of every man/woman in our data set from 2008 to 2016 and age enters the model as controlling variable. 

```{r COX1, echo=FALSE}
###############################
### Exploratory Cox Regression - classic survival setting (time at observation until death or censoring)
###############################

# Preliminary step: Change Reference Category and Category names
link.may_M$cluster2 <- as.factor(link.may_M$cluster2)
link.may_M <- within(link.may_M, cluster2 <- relevel(cluster2, ref = "early onset"))  
link.may_F$cluster2 <- as.factor(link.may_F$cluster2)
link.may_F <- within(link.may_F, cluster2 <- relevel(cluster2, ref = "early onset long"))  

## Males

Cox.CFP.a <- coxph(Surv(time=t.salida,
                        event=event) ~ cluster2 + EDAD, 
                   data=subset(link.may_M))

summary(Cox.CFP.a)


## Females

Cox.CFP.b <- coxph(Surv(time=t.salida,
                        event=event) ~ cluster2 + EDAD, 
                   data=subset(link.may_F))

summary(Cox.CFP.b)
```

Below are the same models with additional covariates (education, civil status, covivienca, last employment situation). I am currently working on a variable to take co-morbidity into account which will add more to the story. We have to be aware of our case numbers ( males = `r length(link.may_F$Id)`, females = `r length(link.may_M$Id)`) but I think it would be better to compare the experience of individuals at the same age with each other. 

```{r COX2, echo=FALSE}

# make them factors
link.may_M$education <- as.factor(link.may_M$education)
link.may_F$education <- as.factor(link.may_F$education)
# change the reference category
link.may_M <- within(link.may_M, education <- relevel(education, ref = "Secondary Educ or higher")) 
link.may_F <- within(link.may_F, education <- relevel(education, ref = "Secondary Educ or higher")) 

# make them factors
link.may_M$civil <- as.factor(link.may_M$civil)
link.may_F$civil <- as.factor(link.may_F$civil)
# change the reference category
link.may_M <- within(link.may_M, civil <- relevel(civil, ref = "Married")) 
link.may_F <- within(link.may_F, civil <- relevel(civil, ref = "Married")) 

# make them factors
link.may_M$work <- as.factor(link.may_M$work)
link.may_F$work <- as.factor(link.may_F$work)
# change the reference category
link.may_M <- within(link.may_M, work <- relevel(work, ref = "Self-employed")) 
link.may_F <- within(link.may_F, work <- relevel(work, ref = "self-employed")) 
# make them factors
link.may_M$CP <- as.factor(link.may_M$CP)
link.may_F$CP <- as.factor(link.may_F$CP)

# change the reference category
link.may_M <- within(link.may_M, CP <- relevel(CP, ref = "Lives with Partner")) 
link.may_F <- within(link.may_F, CP <- relevel(CP, ref = "Lives with Partner"))

link.may_M <- link.may_M %>% mutate(age.gr = as.factor(ifelse(edad5Suecia == "45-64","45-64",
                                             ifelse(edad5Suecia ==  "65-80", "65-80", "81+"))))

link.may_F <- link.may_F %>% mutate(age.gr = as.factor(ifelse(edad5Suecia == "45-64","45-64",
                                                              ifelse(edad5Suecia ==  "65-80", "65-80", "81+"))))

# change the reference category
link.may_M <- within(link.may_M, age.gr <- relevel(age.gr, ref = "45-64")) 
link.may_F <- within(link.may_F, age.gr <- relevel(age.gr, ref = "45-64"))
## Males
## -----

# exit time as time in years (month/12+year)
link.may_M <- link.may_M %>% mutate(t.salida = (a_salida+(m_salida/12)) - 2006.99)
#hist(link.may_M$t.salida)

# entry time (month/12+year) - onset of disability
link.may_M <- link.may_M %>% mutate(t.entrada.dis =  2006.99 - (EDAD-DISCA13_AGE)) # same story with inicio Disca44
link.may_M <- link.may_M %>% mutate(t.salida.dis = (a_salida+(m_salida/12)))

# entry time (month/12+year) - onset of dependency
link.may_M <- link.may_M %>% mutate(t.entrada.dep =  2006.99 - (EDAD-Edadinicio_cuidado)) 

## Females
## -------

# exit time (month/12+year)
link.may_F <- link.may_F %>% mutate(t.salida = (a_salida+(m_salida/12)) - 2006.99)
#hist(link.may_F$t.salida)


# entry time (month/12+year) - onset of disability
link.may_F <- link.may_F %>% mutate(t.entrada.dis =  2006.99 - (EDAD-DISCA13_AGE)) # same story with inicio Disca44
link.may_F <- link.may_F %>% mutate(t.salida.dis = (a_salida+(m_salida/12)))

# entry time (month/12+year) - onset of dependency
link.may_F <- link.may_F %>% mutate(t.entrada.dep =  2006.99 - (EDAD-Edadinicio_cuidado)) 



## Males

Cox.CFP.a <- coxph(Surv(time=t.salida,
                        event=event) ~ cluster2 + EDAD + education + work + CP + civil, 
                   data=subset(link.may_M))

summary(Cox.CFP.a)


## Females

Cox.CFP.b <- coxph(Surv(time=t.salida,
                        event=event) ~ cluster2 + EDAD + education + work + CP + civil, 
                   data=subset(link.may_F))

summary(Cox.CFP.b)
```
One other way to address this issue would be to use age groups instead of a continuous variable. The models could look like this (I will play around with different age groups):
```{r Cox2b, echo=FALSE}
## Males

Cox.CFP.a <- coxph(Surv(time=t.salida,
                        event=event) ~ cluster2 + age.gr + education + work + CP + civil, 
                   data=subset(link.may_M))

summary(Cox.CFP.a)

## Females

Cox.CFP.b <- coxph(Surv(time=t.salida,
                        event=event) ~ cluster2 + age.gr + education + work + CP + civil, 
                   data=subset(link.may_F))

summary(Cox.CFP.b)
```
Below, I ran the models with age at survey as entry variable and age at death/censorship as exit variable. The choice of these time variables might be the most "accurate". I think it may be the most correct way because we compare individuals with the same age with each other and only take the time we really observe into account.
```{r Cox 3, echo=FALSE}
## Males

Cox.CFP.a <- coxph(Surv(time=EDAD,
                        time2 = age.ex,
                        event=event) ~ cluster2 + EDAD, 
                   data=subset(link.may_M))
summary(Cox.CFP.a)


## Females

Cox.CFP.b <- coxph(Surv(time=EDAD,
                        time2 = age.ex,
                        event=event) ~ cluster2 + EDAD, 
                   data=subset(link.may_F))
summary(Cox.CFP.b)

# Plus SES variables Age since onset

# males
Cox.CFP.e <- coxph(Surv(time=EDAD,
                        time2 = age.ex,
                        event=event) ~  cluster2 + EDAD + education + work + CP + civil,
                   data=subset(link.may_M))

summary(Cox.CFP.e)

# females
Cox.CFP.f <- coxph(Surv(time=EDAD,
                        time2 = age.ex,
                        event=event) ~ cluster2 + EDAD + education + work + CP + civil,
                   data=subset(link.may_F))

summary(Cox.CFP.f)
```

Below are the models with the age at dependency as start age (individuals are left-truncated at this age). The results are probably the most interesting, however, I am not 100% sure about this. With this, we select individuals upon there onset age of dependency but theoretically we do not observe there mortality risk unitl the survey year (2008).
```{r Cox4, echo=FALSE}
## Males

Cox.CFP.a <- coxph(Surv(time=Edadinicio_cuidado,
                        time2 = age.ex,
                        event=event) ~ cluster2 + EDAD, 
                   data=subset(link.may_M))
summary(Cox.CFP.a)


## Females

Cox.CFP.b <- coxph(Surv(time=Edadinicio_cuidado,
                        time2 = age.ex,
                        event=event) ~ cluster2 + EDAD, 
                   data=subset(link.may_F))
summary(Cox.CFP.b)


# Plus SES variables Age since onset

# males
Cox.CFP.e <- coxph(Surv(time=Edadinicio_cuidado,
                        time2 = age.ex,
                        event=event) ~  cluster2 + EDAD + education + work + CP + civil,
                   data=subset(link.may_M))

summary(Cox.CFP.e)

# females
Cox.CFP.f <- coxph(Surv(time=Edadinicio_cuidado,
                        time2 = age.ex,
                        event=event) ~ cluster2 + EDAD + education + work + CP + civil,
                   data=subset(link.may_F))

summary(Cox.CFP.f)
```

# Still in Progress (Future Work)

1. Find the right model type - Could be Cox, could be a parametric model
1. Recode the multi-morbity categories
1. Life tables - DFLE
1. Continue with Literature Review - sources about disability in Spain (?)
1. Write everything up - Deadline: 22.03.2019