---
title: "Update 06.03.2019"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(foreign)
library(survival)
library(broom)
library(stargazer)
library(tidyverse)
library(data.table)
# 0.3 Set right working directory
dir()
setwd("C:/Users/y4956294S/Documents/LONGPOP/Subproject 2 - SE differences in transition to dependency/R code/EDAD")

# load traminer data to show graphs
load("datasets/020_traMay_D13.RData")
tra_may_C_50 <- tra_may_D13 %>% filter(EdadInicioDisca13>=50) %>% filter(Edadinicio_cuidado<100) %>% filter(!is.na(EdadInicioDisca13)) %>% 
 filter(DISCA13_AGE<100) %>% filter(EdadInicioDisca13 <= EDAD) 
tra_may_M <- tra_may_C_50 %>% filter(SEXO=="Varón") %>% mutate(SEXO="Male")
tra_may_F <- tra_may_C_50 %>% filter(SEXO=="Mujer") %>% mutate(SEXO="Female")
seqmat_M <- tra_may_M[,c(10:60)]
seqmat_F <- tra_may_F[,c(10:60)]

SeqAlphab_C <- c("DF", "ID", "DC")
# Define color scheme
Brewer_C <- brewer.pal(3, "Set1")

# Males
DisSeq_M <- seqdef(seqmat_M,informat = "STS", alphabet = SeqAlphab_C, id="auto", cpal =  Brewer_C ,start = 55, 
                   labels = c("Disability Free", "Idependent","Care"))

# Females
DisSeq_F <- seqdef(seqmat_F,informat = "STS", alphabet = SeqAlphab_C, id="auto", cpal =  Brewer_C ,start = 55, 
                   labels = c("Disability Free", "Idependent","Care"))

load("datasets/020_traMay_D13M.RData")
load("datasets/020_traMay_D13F.RData")

# Disca 13 - 50+
load(file = 'datasets/030_linkmay_M_50ADL.RData')
load(file = 'datasets/030_linkmay_F_50ADL.RData')

```

# Analysing Sequences to cluster pathways - What is new?
## Clustering
There are a dozens of dissimilarity measures to group individual sequences, many of them more precise than classic OM. Refering to [Studer and Ritschard (2016)](https://rss.onlinelibrary.wiley.com/doi/pdf/10.1111/rssa.12125), I chose to use an OM mechanism sensitive to both the spell length and the timing of sequences (OMspell).


## Population
The population used is a little different for two reasons. First, as I discussed with Antonio, I only used individuals with an onset of disability after age 50 to capture exclusively old-age disability. Second, I found some problems in the matching process that led to unreasonable age distribution (Age of onset of disability after age at survey). As I could not fix all the issues and we still expect the updated data, I deleted the rest of the cases listwise. We end up with `r length(link.may_F$Id)` females and `r length(link.may_M$Id)` male individuals.

## Defining Disability
The onset of disability is defined as age at which the first of the following 13 activities (of daily living) became difficult/was impossible in consequence of a disability or disease.

1. Changing the body posture
1. Walking and moving inside the house
1. Walking and moving outside
1. Sitting down and using public transport
1. Wash and dry different body parts
1. Basic hygene
1. Urination
1. Going to the toilet
1. To dress and undress
1. Eating and drinking
1. Shopping (groceries)
1. Preparing food/cooking
1. Household task (cleaning the house)

# Analysis of mortality and survival
## Descriptive
Since the optimal group/cluster size in all analyses of sequences (choosing different disability indicators) was **two**, we pre-defined two distinct pathways for each sex. See the graphs for men (first graph) and women (second graph).
```{r cluster males, echo=FALSE}
# males 2 and 4 groups
seqdplot(DisSeq_M, group=tra_may_M$cluster2)
seqdplot(DisSeq_F, group=tra_may_F$cluster2)
```
The bars in the following graphs represent mean time in each state by cluster. To keep the same order, men first and then females.
```{r meantime males, echo=FALSE}
# males 2 and 4 groups
seqmtplot(DisSeq_M, group=tra_may_M$cluster2)
seqmtplot(DisSeq_F, group=tra_may_F$cluster2)
```

The following tables are median at death and mean ages at death the men and women separated.
```{r agedeath, echo=FALSE}
# males 2 clusters
link.may_M %>% group_by(cluster2) %>% summarize(mean=mean(age.ex[event==1]), median=median(age.ex[event==1]))
# females 2 cluster
link.may_F %>% group_by(cluster2) %>% summarize(mean=mean(age.ex[event==1]), median=median(age.ex[event==1]))
```
Purely looking at relative distributions, these are durations spent in disability (according to our definition) for all individuals grouped by those who experience the event and those who didn´t.
```{r disdis, echo=FALSE}
# Males
link.may_M <- link.may_M %>% mutate(dur_dis = round(age.ex,0)-round(EdadInicioDisca13,0))
# hist(link.may_M$dur_dis)
summary(link.may_M$dur_dis)
# event distribution by duration in disability
# --------------------------------------------
DUR_Plot_M <- link.may_M %>% mutate(event = as.factor(event)) %>% 
  ggplot(aes(x=dur_dis, fill=event)) +
  geom_histogram(aes(y=0.5*..density..), binwidth=0.5) +
  scale_x_continuous(name = "Duration in Disability (Years)") +
  scale_y_continuous(name = "Relative Frequency",labels = scales::percent) +
  scale_fill_manual(name = "", values=c("#000000", "#A9A9A9"), labels = c("Survivors", "Deceased")) +
  theme_bw()
DUR_Plot_M <- DUR_Plot_M + theme(legend.position = c(0.85, 0.80)) + 
  theme(axis.text=element_text(size=12), axis.title=element_text(size=12,face="bold"))
 
DUR_Plot_M
                                                                          
link.may_F <- link.may_F %>% mutate(dur_dis = round(age.ex,0)-round(EdadInicioDisca13,0))
# hist(link.may_F$dur_dis)
summary(link.may_F$dur_dis)

# event distribution by duration in disability
# --------------------------------------------
DUR_Plot_F <- link.may_F %>% mutate(event = as.factor(event)) %>% 
  ggplot(aes(x=dur_dis, fill=event)) +
  geom_histogram(aes(y=0.5*..density..), binwidth=0.5) +
  scale_x_continuous(name = "Duration in Disability (Years)") +
  scale_y_continuous(name = "Relative Frequency",labels = scales::percent) +
  scale_fill_manual(name = "", values=c("#000000", "#A9A9A9"), labels = c("Survivors", "Deceased")) +
  theme_bw()
DUR_Plot_F <- DUR_Plot_F + theme(legend.position = c(0.85, 0.80)) + theme(axis.text=element_text(size=12),        axis.title=element_text(size=12,face="bold"))

DUR_Plot_F
```

## KME Plots

# Age as time axis
Here we will have an important discussion point. As expected, individuals with a late onset of disability have higher survival probabilities at every single age. Interesting is the difference between men and women (area between the two graphs).
However, these kind of KME don´t make much sense as the curves don´t start at the same point in time and with a risk score lower than 1.

```{r KME1, echo=FALSE}
mfit.1a <- survfit(coxph(Surv(time=EDAD,
                              time2 = age.ex,
                              event = event) ~ 1, data = subset(link.may_M, cluster2=="early onset")), data = subset(link.may_M, cluster2=="early onset"),
                   type = "kaplan-meier")
mfit.1b <- survfit(coxph(Surv(time=EDAD,
                              time2 = age.ex,
                              event = event) ~ 1, data = subset(link.may_M, cluster2=="late onset")), data = subset(link.may_M, cluster2=="late onset"),
                   type = "kaplan-meier")

mfit.1c <- survfit(coxph(Surv(time=EDAD,
                              time2 = age.ex,
                              event = event) ~ 1, data = subset(link.may_F, cluster2=="early onset long")), data = subset(link.may_F, cluster2=="early onset long"),
                   type = "kaplan-meier")

mfit.1d <- survfit(coxph(Surv(time=EDAD,
                              time2 = age.ex,
                              event = event) ~ 1, data = subset(link.may_F, cluster2=="late onset short")), data = subset(link.may_F, cluster2=="late onset short"),
                   type = "kaplan-meier")

KME.Clusta <- tidy(mfit.1a) %>% select(estimate, time) %>% mutate(disab = "early") %>% mutate(sex="male")
KME.Clustb <- tidy(mfit.1b) %>% select(estimate, time) %>% mutate(disab = "late") %>% mutate(sex="male")
KME.Clustc <- tidy(mfit.1c) %>% select(estimate, time) %>% mutate(disab = "early") %>% mutate(sex="female")
KME.Clustd <- tidy(mfit.1d) %>% select(estimate, time) %>% mutate(disab = "late") %>% mutate(sex="female")

KME.CLustM <- union(KME.Clusta, KME.Clustb) %>% union(KME.Clustc) %>% union(KME.Clustd)
KME1 <- KME.CLustM %>% ggplot() +
  geom_step(aes(x=time, y=estimate, color=disab)) +
  scale_y_continuous(name = "Survival Probability")                  +
  scale_x_continuous(name = "Age") +
  scale_colour_manual(values = c("orange", "darkgrey"), name="")     +
  theme_bw()
KME1 + facet_grid(. ~ sex) + theme(legend.position = c(0.85, 0.80)) + theme(axis.text=element_text(size=12),        axis.title=element_text(size=12,face="bold"))
```

Here just for men, the option with more than 2 groups.
```{r KME2, echo=FALSE}

## Males
# exit time (month/12+year)
link.may_M <- link.may_M %>% mutate(t.salida = (a_salida+(m_salida/12)) - 2006.99)
hist(link.may_M$t.salida)
# entry time (month/12+year) - onset of disability
# link.may_M <- link.may_M %>% mutate(t.entrada =  (EDAD-DISCA13_AGE)) # same story with inicio Disca44
# hist(link.may_M$t.entrada)


## Females
# exit time (month/12+year)
link.may_F <- link.may_F %>% mutate(t.salida = (a_salida+(m_salida/12)) - 2006.99)
hist(link.may_F$t.salida)
# entry time (month/12+year) - onset of disability
# link.may_F <- link.may_F %>% mutate(t.entrada = (EDAD-DISCA13_AGE)) # same story with inicio Disca44
# hist(link.may_F$t.entrada)


mfit.2a <- survfit(coxph(Surv(time = t.salida,
                              event = event) ~ 1, data = subset(link.may_M, cluster2=="early onset")), data = subset(link.may_M, cluster2=="early onset"),
                   type = "kaplan-meier")
mfit.2b <- survfit(coxph(Surv(time=t.salida,
                              event = event) ~ 1, data = subset(link.may_M, cluster2=="late onset")), data = subset(link.may_M, cluster2=="late onset"),
                   type = "kaplan-meier")

mfit.2c <- survfit(coxph(Surv(time= t.salida,
                              event = event) ~ 1, data = subset(link.may_F, cluster2=="early onset long")), data = subset(link.may_F, cluster2=="early onset long"),
                   type = "kaplan-meier")

mfit.2d <- survfit(coxph(Surv(time= t.salida,
                              event = event) ~ 1, data = subset(link.may_F, cluster2=="late onset short")), data = subset(link.may_F, cluster2=="late onset short"),
                   type = "kaplan-meier")

KME.Clusta <- tidy(mfit.2a) %>% select(estimate, time) %>% mutate(disab = "early") %>% mutate(sex="male")
KME.Clustb <- tidy(mfit.2b) %>% select(estimate, time) %>% mutate(disab = "late") %>% mutate(sex="male")
KME.Clustc <- tidy(mfit.2c) %>% select(estimate, time) %>% mutate(disab = "early") %>% mutate(sex="female")
KME.Clustd <- tidy(mfit.2d) %>% select(estimate, time) %>% mutate(disab = "late") %>% mutate(sex="female")

KME.CLustM <- union(KME.Clusta, KME.Clustb) %>% union(KME.Clustc) %>% union(KME.Clustd)
KME2 <- KME.CLustM %>% ggplot() +
  geom_step(aes(x=time, y=estimate, color=disab)) +
  scale_y_continuous(name = "Survival Probability")                  +
  scale_x_continuous(name = "Years since survey") +
  scale_colour_manual(values = c("orange", "darkgrey"), name="")     +
  theme_bw()
KME2 + facet_grid(. ~ sex) + theme(legend.position = c(0.85, 0.80)) + theme(axis.text=element_text(size=12),        axis.title=element_text(size=12,face="bold"))

```


## Regression
Due to time pressure, I used Cox PH models to estimate covariate effects on survival. We can discuss using a parametric model (might be more accurate for adult mortality, might be not).

```{r COX1, echo=FALSE}
###############################
### Exploratory Cox Regression
###############################

# Preliminary step: Change Reference Category and Category names
link.may_M$cluster2 <- as.factor(link.may_M$cluster2)
link.may_M <- within(link.may_M, cluster2 <- relevel(cluster2, ref = "late onset"))  
link.may_F$cluster2 <- as.factor(link.may_F$cluster2)
link.may_F <- within(link.may_F, cluster2 <- relevel(cluster2, ref = "late onset"))  

link.may_M$cluster4 <- as.factor(link.may_M$cluster4)
link.may_M <- within(link.may_M, cluster4 <- relevel(cluster4, ref = "late abrupt"))  
link.may_F$cluster3 <- as.factor(link.may_F$cluster3)
link.may_F <- within(link.may_F, cluster3 <- relevel(cluster3, ref = "late abrupt"))  

link.may_M <- within(link.may_M, Education <- relevel(Education, ref = "High"))  
link.may_F <- within(link.may_F, Education <- relevel(Education, ref = "High"))  

link.may_M <- within(link.may_M, Income <- relevel(Income, ref = "625+ eur"))  
link.may_F <- within(link.may_F, Income <- relevel(Income, ref = "625+ eur"))  

# Just the transitions
# males
Cox.CFP.a <- coxph(Surv(time=EDAD,
                        time2=age.ex,
                        event=event) ~ cluster2, 
                   data=subset(link.may_M))

Cox.CFP.a

# females
Cox.CFP.b <- coxph(Surv(time=EDAD,
                        time2=age.ex,
                        event=event) ~ cluster2, 
                   data=subset(link.may_F))

Cox.CFP.b
```

Now the model results with calender time as time axis (entry= Onset year of care).

```{r COX2, echo=FALSE}
# exit time (month/12+year)
link.may_M <- link.may_M %>% mutate(t.salida = a_salida+(m_salida/12))
#hist(link.may_F$t.salida)
# entry time (month/12+year) - onset of disability
link.may_M <- link.may_M %>% mutate(t.entrada = 2006.99 - (EDAD-Edadinicio_cuidado)) # same story with inicio Disca44

# males
Cox.CFP.a <- coxph(Surv(time=t.entrada,
                        time2=t.salida,
                        event=event) ~ cluster2, 
                   data=subset(link.may_M))

Cox.CFP.a

# females
Cox.CFP.b <- coxph(Surv(time=t.entrada,
                        time2=t.salida,
                        event=event) ~ cluster2, 
                   data=subset(link.may_F))

Cox.CFP.b



```

And here the models with the bigger groups:

```{r COX3, echo=FALSE}

# males
Cox.CFP.a <- coxph(Surv(time=t.entrada,
                        time2=t.salida,
                        event=event) ~ cluster4, 
                   data=subset(link.may_M))

Cox.CFP.a

# females
Cox.CFP.b <- coxph(Surv(time=t.entrada,
                        time2=t.salida,
                        event=event) ~ cluster3, 
                   data=subset(link.may_F))

Cox.CFP.b
```

After this overview about how the pathways describe survival in the preliminary models, here with two covariates:

```{r COX4, echo=FALSE}
# Plus SES variables

# males
Cox.CFP.e <- coxph(Surv(time=t.entrada,
                    time2=t.salida,
                    event=event) ~ cluster2 + Education + Income + EDAD,
                 data=subset(link.may_M))

Cox.CFP.e

# females
Cox.CFP.f <- coxph(Surv(time=t.entrada,
                    time2=t.salida,
                    event=event) ~ cluster2 + Education + Income + EDAD,
                 data=subset(link.may_F))

Cox.CFP.f
```

And the, for now, most complete model (information-wise):

```{r COX5, echo=FALSE}
# Plus SES variables

# males
Cox.CFP.e <- coxph(Surv(time=t.entrada,
                    time2=t.salida,
                    event=event) ~ cluster4 + Education + Income + EDAD,
                 data=subset(link.may_M))

summary(Cox.CFP.e)

# females
Cox.CFP.f <- coxph(Surv(time=t.entrada,
                    time2=t.salida,
                    event=event) ~ cluster3 + Education + Income + EDAD,
                 data=subset(link.may_F))

summary(Cox.CFP.f)
```

# Still in Progress (Future Work)

1. Choice of the onset of the pathway - Focus on functional limitation or any disability after age 50?
1. Find the right model - Could be Cox could be a parametric model
1. Clarify the purpose of the analysis
1. Continue with Literature Review - sources about disability in Spain (?)
1. Write everything up - Deadline: 22.03.2019 (1 month)