---
title: "Progress Report - Pathways to Dependency"
author: "Mathias Voigt"
date: "15 de febrero de 2019"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(foreign)
library(survival)
library(broom)
library(stargazer)
```

# General Comments
This document is an attempt to report on the progress and challenges regarding our project on pathways to dependency and their impact on survival disparities. I hope it will help us to stay up-to-date and work efficiently. I tried to address the comments and suggestions from the last meeting and invite you to comment on this document if things are not clear. 
*Suggestion:* We could create a dropbox folder to keep it updated or you send me your comments per email.

## Main Idea
*The general idea will be formulated more eloquently once I started writing on the introduction to the paper.* Up to this point, I would formulate the idea of this project as follows: We are interested in if the health experience (stages of health/functionality) of individuals in need of a personal caretaker can be generalized/grouped. We assume that there are different pathways/trajectories. In a second step, we aim to analyse how such trajectories affect individual mortality risks. I hope you agree with me in this point and I will update this section as soon as possible knowing it is the heart of the pontential paper.
```{r load myData, echo=FALSE, include=FALSE}
setwd("C:/Users/y4956294S/Documents/LONGPOP/Subproject 2 - SE differences in transition to dependency/R code/EDAD")
set.seed(17952)

library(tidyverse)
library(data.table)
library(foreign)
library(survival)
library(broom)
library(stargazer)
library(TraMineR)
library(reshape)
library("RColorBrewer")
library(cluster)
library(WeightedCluster)

load(file='010_mayor.link.RData')
link.may <- link.may %>% filter(LIMIT!="NC") %>% 
  ####
  #### AND at this point only extract the ones who are dependent in 2008
filter(!is.na(Edadinicio_cuidado))

load("020_traMay.RData")
# censored data
load("020_traMay_C.RData")
# censored data with age of onset of a 12 A (ADL group A) disability as transition point from disability free to idependent dis
load("020_traMay_12A.RData")
tra_may_M <- tra_may_C %>% filter(SEXO=="Varón") %>% mutate(SEXO="Male") %>% filter(Edadinicio_cuidado<100)
tra_may_F <- tra_may_C %>% filter(SEXO=="Mujer") %>% mutate(SEXO="Female") %>% filter(Edadinicio_cuidado<100)
tra_may_M_12A <- tra_may_12A %>% filter(SEXO=="Varón") %>% mutate(SEXO="Male") %>% filter(Edadinicio_cuidado<100)
tra_may_F_12A <- tra_may_12A %>% filter(SEXO=="Mujer") %>% mutate(SEXO="Female") %>% filter(Edadinicio_cuidado<100)
seqmat_M <- tra_may_M[,c(14:64)]
seqmat_M_12A <- tra_may_M_12A[,c(14:64)]
seqmat_F <- tra_may_F[,c(14:64)]
seqmat_F_12A <- tra_may_F_12A[,c(14:64)]
# Alphabet
SeqAlphab_C <- c("DF","ID", "DC")
# Define color scheme
Brewer_C <- brewer.pal(3, "Set1")
DisSeq_M <- seqdef(seqmat_M,informat = "STS", alphabet = SeqAlphab_C, id="auto", cpal =  Brewer_C ,start = 55, 
                   labels = c("Disability Free","Idependent","Care"))
DisSeq_M12A <- seqdef(seqmat_M_12A,informat = "STS", alphabet = SeqAlphab_C, id="auto", cpal =  Brewer_C ,start = 55, 
                   labels = c("Disability Free","Idependent","Care"))
DisSeq_F <- seqdef(seqmat_F,informat = "STS", alphabet = SeqAlphab_C, id="auto", cpal =  Brewer_C ,start = 55, 
                   labels = c("Disability Free","Idependent","Care"))
DisSeq_F12A <- seqdef(seqmat_F_12A,informat = "STS", alphabet = SeqAlphab_C, id="auto", cpal =  Brewer_C ,start = 55, 
                   labels = c("Disability Free","Idependent","Care"))
```
## Population
There are two subpopulations of the EDAD survey that differ slightly due to missing information on central variables. The base population includes all individuals who are older than 65 years at the time of observation in 2008 (EDAD survey), linked to MNP death register or Padron, and have stated to receive some form of personal care. In one case I used the population who stated that they have started to encounter problems with activities of independent daily life (IADL - grupo A). In this case I had to exclude a few people with missing information on these variables. I also excluded outlier with an onset age of disability older than 100.

# Variables of interest
*This will be updated soon*

# Age at onset of dependency
The descriptive data suggest that there are cases of age heaping especially for the age of onset of disability (probably due to recall errors). The graph below depicts the distribution of ages at onset of dependency by severity (2 categories). Antonio mentioned that age information may be biased as they asked elderly/disabled or their caretakers about a large number of different limitations.  We can discuss smoothing and redistribution of ages of onset of disabilty/dependency.
```{r depend, echo=FALSE}
summary(link.may$Edadinicio_cuidado)
link.may %>% 
  ggplot(aes(x=Edadinicio_cuidado, fill=Gravity)) +
  geom_histogram(binwidth =  1) +
  scale_x_continuous(name = "Age") +
  scale_fill_discrete(name = "") +
  theme_bw()
```
## Age at onset of disability (all 44 disabilities):
```{r disabil, echo=FALSE}
summary(link.may$EdadInicioDisca44)
link.may %>% 
  ggplot(aes(x=EdadInicioDisca44, fill=Gravity)) +
  geom_histogram(binwidth =  5) +
  scale_x_continuous(name = "Age") +
  scale_fill_discrete(name = "") +
  theme_bw()
```

```{r edadmuerte, echo=FALSE}
# The distribution of ages at death or right censoring:
# link.may <- link.may %>% mutate(event = as.factor(ifelse(event==1, "dead", "alive")))
# 
# link.may %>%
#   ggplot(aes(x=age.ex, fill=event)) +
#   geom_histogram(bins = 44) +
#   scale_x_continuous(name = "Age") +
#   scale_fill_discrete(name = "") +
#   theme_bw()

# 
# link.may <- data.table(link.may)
# with(link.may, tapply(age.ex, list(event,sex), mean))
```
# Clustering
I played around with different techniques over the last two weeks and will give you the data and plots on the clustering mechanism that, in my opinion, suits our analysis best. Please let me know if you have questions. First, I decided to treat censoring as independent in contrast of treating it as a state as before. That means that an individual sequence stops at the last year of information (age of the individual in 2008). From there on I used two different onset variables to define if someone is disability-free or lives independently with disability. In case A.), I took the onset age for any disability (all 44 variables) for defining the first state change from disability-free to living with disability. In the other case (B.) ), I used the onset age when someone started to have problems with activities of independent living (type A functionality disorder = Onset of a IADL).
## Cluster- or group selection with all disabilities (case A.))
These are the ordered state sequences for males (left) and females (right) for case A.). The color legend has changed a little bit (red= disability free, blue= disabled but "independent", and green= dependent on personal care taker).

```{r cluster, echo=FALSE}
# 1.4.1 Sorted sequences for both sexes
par(mfrow=c(1,2))
seqplot(DisSeq_M,type = "I", with.legend = FALSE, sort="from.end")
seqplot(DisSeq_F,type = "I", with.legend = FALSE, sort="from.end")
```
Below is the state distribution plot showing the relative distribution of states by age (values for later ages are distorted due to small case numbers = relative distribution). The other two plots show the modal state at any given age.

```{r dplot, echo=FALSE}
par(mfrow=c(1,3))
seqdplot(DisSeq_M, with.legend = FALSE)
seqdplot(DisSeq_F, with.legend = FALSE)
seqlegend(DisSeq_M)
par(mfrow=c(1,1))
seqmsplot(DisSeq_M)
seqmsplot(DisSeq_F)
```
From these state sequence dataset we create a cost matrix which will help us to group the sequences. The last time I showed you results based on the classic OM approach. This method seems to be outdated so, I chose a adjusted OM process that takes the sequence lenght into account (that was one point in the meeting). There are many other methods which work better for different kind of data or interests (sequence lengths, sequence order etc.).
```{r clustermatching, echo=FALSE, include=FALSE}
# males
submat_M <- seqsubm(DisSeq_M, method = "CONSTANT", cval = 2)
# females
submat_F <- seqsubm(DisSeq_F, method = "CONSTANT", cval = 2)
# males (OMslen - spell-length sensitive)
dismat_M2 <- seqdist(DisSeq_M, method = "OMslen", sm = submat_M)
clusterw_M2 <- agnes(dismat_M2, diss = TRUE, method = "ward")

# females (OMslen - spell-length sensitive)
dismat_F2 <- seqdist(DisSeq_F, method = "OMslen", sm = submat_F)
clusterw_F2 <- agnes(dismat_F2, diss = TRUE, method = "ward")
```
There is a statistical test to obtain the best number of clusters/groups. If you wish, I can forward you the literature. The test statistic that is suggested to pay attention to is the ASW. The first suggestion in both cases seem to be 2 groups. To account for a little bit more heterogeneity, I would consider the second suggestion as well which is 4 groups for men and 3 for women.
```{r teststat, echo=FALSE}
data.clust_M2 <- wcKMedRange(dismat_M2, kvals=2:20)
summary(data.clust_M2, max.rank = 2)
data.clust_F2 <- wcKMedRange(dismat_F2, kvals=2:20)
summary(data.clust_F2, max.rank = 2)
cluster2M2 <- cutree(clusterw_M2, k = 2)
cluster4M2 <- cutree(clusterw_M2, k = 4)
cluster2F2 <- cutree(clusterw_F2, k = 2)
cluster3F2 <- cutree(clusterw_F2, k = 3)

cluster2M2 <- factor(cluster2M2, labels = c("Type 1", "Type 2"))
cluster4M2 <- factor(cluster4M2, labels = c("Type 1", "Type 2", "Type 3","Type 4"))
cluster2F2 <- factor(cluster2F2, labels = c("Type 1", "Type 2"))
cluster3F2 <- factor(cluster3F2, labels = c("Type 1", "Type 2", "Type 3"))

```
In the first graph below there are example sequences for each group/cluster similar to those I showed you at the last meeting. The second type of graph shows the mean time spent in each state by group. The first 4 graphs are regarding males, the second 4 females. While in the two the version with the 2 clusters we just distinguish between the ones with long periods of disability before receiving care and those with very rapid transitions, the 4 scenario appears to give us a little bit more variation about when things happen (earlier or later in life). Personally, I think that could add an important piece of information.
```{r clustergraphs, echo=FALSE}
seqfplot(DisSeq_M, group = cluster2M2, pbarw = F)
seqfplot(DisSeq_M, group = cluster4M2, pbarw = F)
seqmtplot(DisSeq_M, group = cluster2M2)
seqmtplot(DisSeq_M, group = cluster4M2)

seqfplot(DisSeq_F, group = cluster2F2, pbarw = F)
seqfplot(DisSeq_F, group = cluster3F2, pbarw = F)
seqmtplot(DisSeq_F, group = cluster2F2)
seqmtplot(DisSeq_F, group = cluster3F2)

# here the part with the cluster assigning back to the data
# males 
# -----

# 2 groups (OMslen object)
tra_may_M$clusters2 <- data.clust_M2$clustering$cluster2

# 4 groups (OMslen object)
tra_may_M$clusters4 <- data.clust_M2$clustering$cluster4

# females 
# -------

# 2 groups (OMslen object)
tra_may_F$clusters2 <- data.clust_F2$clustering$cluster2

# 4 groups (OMslen object)
tra_may_F$clusters3 <- data.clust_F2$clustering$cluster3

# males
tra_may_M <- tra_may_M %>% select(Id, clusters2, clusters4) %>% 
              mutate(cluster2 = ifelse(clusters2==177, "late onset", "early onset")) %>% 
              mutate(cluster4 = ifelse(clusters4==143, "late gradual", ifelse(clusters4==1472, "early gradual",
                                 ifelse(clusters4==256, "late healthy", "late abrupt")))) %>% 
              select(-clusters4, - clusters2)

# females
tra_may_F <- tra_may_F %>% select(Id, clusters2, clusters3) %>%
             mutate(cluster2 = ifelse(clusters2==1844, "late onset short", "early onset long")) %>%
             mutate(cluster3 = ifelse(clusters3==1214, "early abrupt", 
                    ifelse(clusters3==1426,"late abrupt", "early gradual"))) %>% 
             select(-clusters2, -clusters3)
# males
tra_may_M <- tra_may_M[order(tra_may_M$Id),]

# males
tra_may_F <- tra_may_F[order(tra_may_F$Id),]

## 3.3 Link back to big data set!

# order big data + extract the two women with late onset of anything

link.may <- link.may %>% filter(Edadinicio_cuidado<100)

# make 2 data sets out of them (by sex) and order them

link.may_M <- link.may %>% filter(Sex=="Male")
link.may_F <- link.may %>% filter(Sex=="Female")


link.may_M <- link.may_M[order(link.may_M$Id),]
link.may_F <- link.may_F[order(link.may_F$Id),]

## 3.4  Create working data sets for the mortality analysis
link.may_M <- link.may_M %>% left_join(tra_may_M, by="Id")
link.may_F <- link.may_F %>% left_join(tra_may_F, by="Id")

```
Below I estimated the median age at death (after age 65) for each group. I do not see anything extremely surprising. Looks like a late onset of disability is associated with at later average age at death. Of course that will be double-checked and tested against covariates.
```{r median age at death by group, echo=FALSE}
# males 2 clusters
link.may_M %>% mutate(Male2Groups=cluster2) %>% group_by(Male2Groups) %>% summarize(mean=mean(age.ex[event==1]), median=median(age.ex[event==1]))
# males 3 clusters
link.may_M %>% mutate(Male4Groups=cluster4) %>% group_by(Male4Groups) %>% summarize(mean=mean(age.ex[event==1]), median=median(age.ex[event==1]))

# females 2 clusters
link.may_F %>% mutate(Female2Groups=cluster2) %>% group_by(Female2Groups) %>% summarize(mean=mean(age.ex[event==1]), median=median(age.ex[event==1]))
# females 3 clusters
link.may_F %>% mutate(Female3Groups=cluster3) %>% group_by(Female3Groups) %>% summarize(mean=mean(age.ex[event==1]), median=median(age.ex[event==1]))

```
## Cluster- or group selection by functionality - IADL (case B.))
From here on I will go over the same procedure as before but use a different selection criteria for the onset of disability The dataset is a little bit smaller because there were a handful of females for which we did not have the information.
```{r cluster12A, echo=FALSE}
# 1.4.1 Sorted sequences for both sexes
par(mfrow=c(1,2))
seqplot(DisSeq_M12A,type = "I", with.legend = FALSE, sort="from.end")
seqplot(DisSeq_F12A,type = "I", with.legend = FALSE, sort="from.end")

par(mfrow=c(1,3))
seqdplot(DisSeq_M12A, with.legend = FALSE)
seqdplot(DisSeq_F12A, with.legend = FALSE)
seqlegend(DisSeq_M12A)
par(mfrow=c(1,1))
seqmsplot(DisSeq_M12A)
seqmsplot(DisSeq_F12A)

# males
submat_MA12 <- seqsubm(DisSeq_M12A, method = "CONSTANT", cval = 2)
# females
submat_FA12 <- seqsubm(DisSeq_F12A, method = "CONSTANT", cval = 2)
# males (OMslen - spell-length sensitive)
dismat_M2A12 <- seqdist(DisSeq_M12A, method = "OMslen", sm = submat_MA12)
clusterw_M2A12 <- agnes(dismat_M2A12, diss = TRUE, method = "ward")

# females (OMslen - spell-length sensitive)
dismat_F2A12 <- seqdist(DisSeq_F12A, method = "OMslen", sm = submat_FA12)
clusterw_F2A12 <- agnes(dismat_F2A12, diss = TRUE, method = "ward")
```
Interesting enough, these graphs seem to tell a story closer to how I imagine the reality, at least from the distributions at the older ages and the modal state, which is "receiving care" after a certain age. In case B), while the best scenario seems to be 2 groups for both sexes, the second best group number is 3 for males and 5 for females. As in case A) I went on with both group/cluster sizes.
```{r teststat2, echo=FALSE}
data.clust_M2A12 <- wcKMedRange(dismat_M2A12, kvals=2:20)
summary(data.clust_M2A12, max.rank = 2)
data.clust_F2A12 <- wcKMedRange(dismat_F2A12, kvals=2:20)
summary(data.clust_F2A12, max.rank = 2)
# the best two options
cluster2MA12_2 <- cutree(clusterw_M2A12, k = 2)
cluster3MA12_2 <- cutree(clusterw_M2A12, k = 3)
cluster2FA12_2 <- cutree(clusterw_F2A12, k = 2)
cluster5FA12_2 <- cutree(clusterw_F2A12, k = 5)

cluster2MA12_2 <- factor(cluster2MA12_2, labels = c("Type 1", "Type 2"))
cluster3MA12_2  <- factor(cluster3MA12_2, labels = c("Type 1", "Type 2", "Type 3"))
cluster2FA12_2 <- factor(cluster2FA12_2, labels = c("Type 1", "Type 2"))
cluster5FA12_2 <- factor(cluster5FA12_2, labels = c("Type 1", "Type 2", "Type 3","Type 4", "Type 5"))

```
```{r clustergraphs2, echo=FALSE}
seqfplot(DisSeq_M12A, group = cluster2MA12_2, pbarw = F)
seqfplot(DisSeq_M12A, group = cluster3MA12_2, pbarw = F)
seqmtplot(DisSeq_M12A, group = cluster2MA12_2)
seqmtplot(DisSeq_M12A, group = cluster3MA12_2)

seqfplot(DisSeq_F12A, group = cluster2FA12_2, pbarw = F)
seqfplot(DisSeq_F12A, group = cluster5FA12_2, pbarw = F)
seqmtplot(DisSeq_F12A, group = cluster2FA12_2)
seqmtplot(DisSeq_F12A, group = cluster5FA12_2)

# here the part with the cluster assigning back to the data
# males 
# -----

# 2 groups (OMslen object)
tra_may_M_12A$clusters2 <- data.clust_M2A12$clustering$cluster2

# 4 groups (OMslen object)
tra_may_M_12A$clusters3 <- data.clust_M2A12$clustering$cluster3

# females 
# -------

# 2 groups (OMslen object)
tra_may_F_12A$clusters2 <- data.clust_F2A12$clustering$cluster2

# 4 groups (OMslen object)
tra_may_F_12A$clusters5 <- data.clust_F2A12$clustering$cluster5

# males
tra_may_M_12A <- tra_may_M_12A %>% select(Id, clusters2, clusters3) %>%
              mutate(cluster2 = ifelse(clusters2==5, "late onset", "early onset")) %>%
              mutate(cluster3 = ifelse(clusters3==345, "late gradual", ifelse(clusters3==88, "early gradual","late abrupt"))) %>%           select(-clusters2, - clusters3)

# females
tra_may_F_12A <- tra_may_F_12A %>% select(Id, clusters2, clusters5) %>%
             mutate(cluster2 = ifelse(clusters2==281, "late onset short", "early onset long")) %>%
             mutate(cluster5 = ifelse(clusters5==3001, "early abrupt",
                    ifelse(clusters5==56,"late abrupt", ifelse(clusters5==1852,"middle gradual", 
                                                               ifelse(clusters5==3018, "early gradual", "late gradual"))))) %>%
             select(-clusters2, -clusters5)
# males
tra_may_M_12A <- tra_may_M_12A[order(tra_may_M_12A$Id),]

# males
tra_may_F_12A <- tra_may_F_12A[order(tra_may_F_12A$Id),]

# order big data + extract the two women with late onset of anything

link.may12A <- link.may %>% filter(Edadinicio_cuidado<100) %>% filter(!is.na(edadiniciodisca12A))

# make 2 data sets out of them (by sex) and order them

link.may_M12A <- link.may12A %>% filter(Sex=="Male")
link.may_F12A <- link.may12A %>% filter(Sex=="Female")

## 3.4  Create working data sets for the mortality analysis
link.may_M12A <- link.may_M12A %>% left_join(tra_may_M_12A, by="Id")
link.may_F12A <- link.may_F12A %>% left_join(tra_may_F_12A, by="Id")

```
And again the median and mean ages at death. This is of course just a descriptive statistic but the differences appear to be much smaller and are therefore harder to interpret.
```{r median age at death by group2, echo=FALSE}
# males 2 clusters
link.may_M12A %>% mutate(Male2Groups=cluster2) %>% group_by(Male2Groups) %>% summarize(mean=mean(age.ex[event==1]), median=median(age.ex[event==1]))
# males 3 clusters
link.may_M12A %>% mutate(Male3Groups=cluster3) %>% group_by(Male3Groups) %>% summarize(mean=mean(age.ex[event==1]), median=median(age.ex[event==1]))

# females 2 clusters
link.may_F12A %>% mutate(Female2Groups=cluster2) %>% group_by(Female2Groups) %>% summarize(mean=mean(age.ex[event==1]), median=median(age.ex[event==1]))
# females 3 clusters
link.may_F12A %>% mutate(Female5Groups=cluster5) %>% group_by(Female5Groups) %>% summarize(mean=mean(age.ex[event==1]), median=median(age.ex[event==1]))

```